---
title: "BMEG 310 Assignment 2"
author: "Group 14: Wesley Chan, Flora Deng, Mary Graves"
date: "2022-10-17"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# loading in data
ovarian.dataset <- read.delim("ovarian.data", sep=",", header = FALSE)
features <- c("perimeter", "area", "smoothness", "symmetry", "concavity",
paste("protein", seq(1, 25) ))
names(ovarian.dataset) <- c("cell_id", "diagnosis", features)

# check data (REMOVE FOR SUBMISSION)
head(ovarian.dataset)
```

\newpage

# Question 1: Dimensionality Reduction

### Q1.1

```{r}
ovarian.pca <- prcomp(ovarian.dataset[,c(3:32)], center = TRUE,scale. = TRUE)

summary(ovarian.pca)
```

According to the proportion of variance in summary, 44.3% variation is associated with PC1.


### Q1.2

```{r}
screeplot(ovarian.pca, type = "l", npcs = 30, main = "Screeplot of the 30 PCs")
abline(h = 1, col="red")
legend("topright", legend=c("Eigenvalue = 1"),
       col=c("red"))

cumpro <- cumsum(ovarian.pca$sdev^2 / sum(ovarian.pca$sdev^2))
plot(cumpro[0:30], xlab = "PC #", ylab = "Amount of explained variance", main = "Cumulative variance plot")
abline(v = 6, col="blue")
abline(h = 0.88759, col="blue")
legend("topleft", legend=c("Cut-off @ PC6"),
       col=c("blue"))
```

Based on cumulative proportion we get in Q1.1 and plots we get, we can see that the first 6 components (PC1-PC7) has an eigenvalue > 1 and represent 90% of the variance in the data. We can preserve 90% of the variability by 6 dimensionality of the reduced feature space.

### Q1.3

```{r warning=FALSE}
library(devtools)
remotes::install_github("vqv/ggbiplot")
library(ggbiplot)

ggbiplot(ovarian.pca, labels=ovarian.dataset$diagnosis ,groups = ovarian.dataset$diagnosis)
```

### Q1.4

```{r warning=FALSE}
library(ggplot2)
ggplot(data = ovarian.dataset, 
       mapping = aes(x = area, y = concavity, color = as.factor(diagnosis)))+ 
       geom_point()
```

### Q1.5
I think the first figure is better because the distribution of points in Figure 1 is more dense and there is not much overlap. 

### Q1.6

```{r}
boxplot(ovarian.pca$x)
```

\newpage

# Question 2 - Clustering

### Q2.1

```{r}
library(FactoMineR)
library(factoextra)

kmeans.data <- as.data.frame(scale(ovarian.dataset[,c(3:32)]))

set.seed(43827)
kmeans.results = kmeans(kmeans.data, centers=2, nstart=50)
results = table(kmeans.results$cluster, ovarian.dataset$diagnosis)

if (results[1] < results[2]) {
  labels = ifelse(kmeans.results$cluster == 1, "M", "B")
} else {
  labels = ifelse(kmeans.results$cluster == 1, "B", "M")
}

cm = table(labels, ovarian.dataset$diagnosis)

accuracy = (cm[1]+cm[4])/(cm[1]+cm[2]+cm[3]+cm[4])
precision = (cm[1])/(cm[1]+cm[2])
recall = (cm[1])/(cm[1]+cm[3])

cat("Accuracy:", accuracy, "\n")
cat("Precision:", precision, "\n")
cat("Recall:", recall, "\n")
```

### Q2.2 
```{r}
accuracies = replicate(10,0)

set.seed(43827)
for (i in 1:10) {
  kmeans.results = kmeans(kmeans.data, centers=2, nstart=1)
  results = table(kmeans.results$cluster, ovarian.dataset$diagnosis)

  if (results[1] < results[2]) {
    labels = ifelse(kmeans.results$cluster == 1, "M", "B")
  } else {
    labels = ifelse(kmeans.results$cluster == 1, "B", "M")
  }

  cm = table(labels, ovarian.dataset$diagnosis)
  
  accuracy = (cm[1]+cm[4])/(cm[1]+cm[2]+cm[3]+cm[4])
  accuracies[i] = accuracy
}

cat("Accuracies of Runs:", accuracies, "\n")
cat("Mean of Accuracies:", mean(accuracies), "\n")
```

When running kmeans clustering for this question, we set nstart=1 in the kmeans function. This causes the results to be slightly different each run, because the start points for clustering are random (and thus the clusters might be slightly different). In Q2.1, we set nstart=50, which causes R to run the clustering 50 times with random starting assignments, then select the best run based on lowest within-cluster variation. This helps alleviate the randomness in the results and give us more stable clusters, although it takes longer to run. 

### Q2.3
For this question, we will use the method in Q2.2, with nstart=1, since we are comparing the results from this question with Q2.2. 

```{r}
kmeans_top5pca <- ovarian.pca$x[,1:5]
kmeans.data.pca <- data.frame(cbind(kmeans_top5pca, ovarian.dataset$diagnosis))[,c(1:5)]

set.seed(33422) #want a different seed for this analysis
kmeans.results.pca = kmeans(kmeans.data.pca, centers=2)
results.pca = table(kmeans.results.pca$cluster, ovarian.dataset$diagnosis)

if (results.pca[1] < results.pca[2]) {
  labels.pca = ifelse(kmeans.results.pca$cluster == 1, "M", "B")
} else {
  labels.pca = ifelse(kmeans.results.pca$cluster == 1, "B", "M")
}

cm.pca = table(labels.pca, ovarian.dataset$diagnosis)

accuracy.pca = (cm.pca[1]+cm.pca[4])/(cm.pca[1]+cm.pca[2]+cm.pca[3]+cm.pca[4])
precision.pca = (cm.pca[1])/(cm.pca[1]+cm.pca[2])
recall.pca = (cm.pca[1])/(cm.pca[1]+cm.pca[3])

cat("Accuracy:", accuracy.pca, "\n")
cat("Precision:", precision.pca, "\n")
cat("Recall:", recall.pca, "\n")

```

### Q2.4
Between Q2.2 and Q2.3, we can see that kmeans clustering with the first five PCs has a slightly lower accuracy, since we are missing some information from the other PCs.

\newpage

# Question 3 - Classification 

```{r}
# Divide data into training set and test set
ovarian.dataset.train <- ovarian.dataset[sample(nrow(ovarian.dataset))
  [1:(nrow(ovarian.dataset)/2)],]
ovarian.dataset.test <- ovarian.dataset[sample(nrow(ovarian.dataset))
  [(nrow(ovarian.dataset)/2):(nrow(ovarian.dataset))],]
```

### Q3.1

```{r}
glm.fit <- glm(as.factor(diagnosis)~., data = ovarian.dataset.train[,c(3:32)], 
               family = binomial)

summary(glm.fit)
```

```{r}
glm.probs <- predict(glm.fit, ovarian.dataset.test[,c(3:32)],
                     type = "response")

glm.probs[1:5]
```

```{r}
glm.pred <- ifelse(glm.probs > 0.5, "M", "B")
head(ovarian.dataset.test)
table(glm.pred,ovarian.dataset.test$diagnosis)
mean(glm.pred == ovarian.dataset.test$diagnosis)
```

### Q3.2

```{r}
ovarian.pca5 <- ovarian.pca$x[,1:5]
ovarian.dataset.train.pc5 <- ovarian.pca5[sample(nrow(ovarian.pca5))[1:(nrow(ovarian.pca5)/2)],]
ovarian.dataset.test.pc5 <- ovarian.pca5[sample(nrow(ovarian.pca5))[(nrow(ovarian.pca5)/2):(nrow(ovarian.pca5))],]

glm.fit.pc5 <- glm(as.factor(diagnosis) ~., data = as.data.frame(ovarian.dataset.train.pc5), family = binomial)

summary(glm.fit.pc5)
```

```{r}
glm.probs.pc5 <- predict(glm.fit.pc5,as.data.frame(ovarian.dataset.test.pc5),type = "response")

glm.probs.pc5[1:5]
```

```{r}
glm.pred.pc5 <- ifelse(glm.probs.pc5 > 0.5, "M", "B")
head(as.data.frame(ovarian.dataset.test.pc5))
table(glm.pred.pc5,ovarian.dataset.test$diagnosis)
mean(glm.pred.pc5 == ovarian.dataset.test$diagnosis)
```


### Q3.3
The results in Q3.2 is better since the accuracy is larger. Top 5 PCs have greater association variance.


### Q3.4

```{r}
# gotta do
```

### Q3.5

```{r}
library(ROCR)
pred.prob <- predict(glm.fit, ovarian.dataset, type="response")
predict <- prediction(pred.prob, ovarian.dataset$diagnosis, label.ordering=c("B","M"))
perform <- performance(predict,"tpr","fpr")
plot(perform,colorize=TRUE)
```

### Q3.6

```{r}
library(randomForest)

set.seed(123)

chunk <- sample(nrow(ovarian.dataset), 0.5 * nrow(ovarian.dataset))

training_dataset <- ovarian.dataset[chunk, ]

testing_dataset <- ovarian.dataset[-chunk, ]
```

```{r}
# Question 3.6 - (3.1)
model <- randomForest(as.factor(diagnosis) ~ ., training_dataset)
summary(model)
```

```{r}
prediction <- predict(model, newdata = testing_dataset)
prediction[1:5]
table(prediction,testing_dataset$diagnosis)
mean(prediction == testing_dataset$diagnosis)
```

```{r}
# Question 3.6 - (3.2)
training_dataset.pca <- prcomp(training_dataset[,c(3:32)], center = TRUE,scale. = TRUE)
testing_dataset.pca <- prcomp(testing_dataset[,c(3:32)], center = TRUE,scale. = TRUE)

training_dataset.pc5 <- training_dataset.pca$x[,1:5]
testing_dataset.pc5 <- testing_dataset.pca$x[,1:5]

model.testing_dataset.pc5 <- randomForest(as.factor(testing_dataset$diagnosis) ~ ., testing_dataset.pc5)
summary(model.testing_dataset.pc5)
```

```{r}
prediction.pca <- predict(model.testing_dataset.pc5, newdata = testing_dataset.pc5)
prediction.pca[1:5]
table(prediction.pca,testing_dataset$diagnosis)
mean(prediction.pca == testing_dataset$diagnosis)
```

\newpage

# Contribution Statement

Wesley Chan (75520023): did literally nothing
Flora Deng (): 
Mary Graves (): 
